skater: An R package for SNP-based Kinship Analysis, Testing, and Evaluation

### 2021-07-14 paper/paper.tex
update supplement and recompile

Inferring familial relationships between individuals using genetic data
is a common problem in population genetics, medical genetics, and
forensics. There are multiple approaches to estimating relatedness
between samples, including genome-wide measures such as those
implemented in Plink \citep{purcell2007} or KING
\citep{manichaikul2010}, and methods that rely on identity by descent
(IBD) segment detection such as GERMLINE \citep{gusev2009}, hap-IBD
\citep{zhou2020}, and IBIS \citep{seidman2020}. Recent efforts focusing
on benchmarking these methods \citep{ramstetter2017, vries2021} have
been aided by tools for simulating pedigrees and genome-wide SNP data
\citep{caballero2019}. Analyzing results from genome-wide SNP-based
kinship analysis or comparing analyses to simulated data for
benchmarking have to this point required writing one-off analysis
functions or utility scripts that are seldom distributed with robust
documentation, test suites, or narrative examples of usage. There is a
need in the field for a well-documented software package with a
consistent design and API that contains functions to assist with
downstream manipulation, benchmarking, and analysis of SNP-based kinship
assessment methods. Here we present the skater package for
\textbf{S}NP-based \textbf{k}inship \textbf{a}nalysis, \textbf{t}esting,
and \textbf{e}valuation with \textbf{R}.

@@

The \texttt{dibble()} function creates a \textbf{d}egree
\textbf{i}nference \textbf{t}ibble, with degrees up to the specified
maximum degree resolution, expected kinship coefficient, and lower and
upper inference ranges as defined in \citet{manichaikul2010}. The
\texttt{kin2degree()} function infers the relationship degree given a
kinship coefficient and a maximum degree resolution (e.g., 7th-degree
relatives) up to which anything more distant is treated as unrelated.

Once estimated kinship is converted to degree, it may be of interest to
compare the inferred degree to truth. When aggregated over many
relationships and inferences, this can help benchmark performance of a
particular kinship analysis method. The skater package adapts a
\texttt{confusion\_matrix()} function from \citet{clark2021} to provide
standard contingency table metrics (e.g.~sensitivity, specificity, PPV,
precision, recall, F1, etc.) with a new reciprocal RMSE (R-RMSE) metric.
The R-RMSE metric is defined more thoroughly in the skater package
vignette (see Supplementary Material), and may be a preferrable measure
of classification accuracy when benchmarking relationship degree
estimation. Taking the reciprocal of the target and predicted degree in
a typical RMSE calculation results in larger penalties for more
egregious misclassifications (e.g., classifying a first-degree relative
pair as second degree) than misclassifications at more distant
relationships (e.g., classifying a fourth-degree relative pair as
fifth-degree).

>>>>

Inferring familial relationships between individuals using genetic data
is a common practice in population genetics, medical genetics, and
forensics. There are multiple approaches to estimating relatedness
between samples, including genome-wide measures such as those
implemented in Plink \citep{purcell2007} or KING
\citep{manichaikul2010}, and methods that rely on identity by descent
(IBD) segment detection such as GERMLINE \citep{gusev2009}, hap-IBD
\citep{zhou2020}, and IBIS \citep{seidman2020}. Recent efforts focusing
on benchmarking these methods \citep{ramstetter2017, vries2021} have
been aided by tools for simulating pedigrees and genome-wide SNP data
\citep{caballero2019}. Analyzing results from genome-wide SNP-based
kinship analysis or comparing analyses to simulated data for
benchmarking have to this point required writing one-off analysis
functions or utility scripts that are seldom distributed with robust
documentation, test suites, or narrative examples of usage. There is a
need in the field for a well-documented software package with a
consistent design and API that contains functions to assist with
downstream manipulation, benchmarking, and analysis of SNP-based kinship
assessment methods. Here we present the skater package for
\textbf{S}NP-based \textbf{k}inship \textbf{a}nalysis, \textbf{t}esting,
and \textbf{e}valuation with \textbf{R}.

@@

The \texttt{dibble()} function creates a \textbf{d}egree
\textbf{i}nference t\textbf{ibble}, with degrees up to the specified
maximum degree resolution, expected kinship coefficient, and lower and
upper inference ranges as defined in \citet{manichaikul2010}. The
\texttt{kin2degree()} function infers the relationship degree given a
kinship coefficient and a maximum degree resolution (e.g., 7th-degree
relatives) up to which anything more distant is classified as unrelated.

Once estimated kinship is converted to degree, it may be of interest to
compare the inferred degree to known degrees of relatedness. When
aggregated over many relationships and inferences, this can help
benchmark performance of a particular kinship analysis method. The
skater package adapts a \texttt{confusion\_matrix()} function from
\citet{clark2021} to provide standard contingency table metrics
(e.g.~sensitivity, specificity, PPV, precision, recall, F1, etc.) with a
new reciprocal RMSE (R-RMSE) metric. The R-RMSE metric is defined more
thoroughly in the skater package vignette (see Supplementary Material),
and may be a preferrable measure of classification accuracy when
benchmarking relationship degree estimation. Taking the reciprocal of
the target and predicted degree in a typical RMSE calculation results in
larger penalties for more egregious misclassifications (e.g.,
classifying a first-degree relative pair as second degree) than
misclassifications at more distant relationships (e.g., classifying a
fourth-degree relative pair as fifth-degree).
